---
title: "Predictive Forecast for COVID-19 situation in Canada"
author: "Peter Hou"
date: "03/05/2020"
output:
  pdf_document: defaultp
  html_document: default
---


# Objective:
The last report focusing on the specific number of cases occured at different regions. This project uses data where the detailed condition of 1085 COVID infected patients. 

The major objective of this product is to create model for classifying any given COVID-19 patients with a set of important information about them as likely to recover or no.

``` {r}
COVID_feature <- read.csv("COVID19_line_list_data.csv")
summary(COVID_feature)
```


# Date Cleaning
``` {r}
cleaned_COVID_feature <- COVID_feature[c("gender", "age",
 "symptom","death","recovered")]
## Remove NA gender column
cleaned_COVID_feature <-
 cleaned_COVID_feature[which(!is.na(cleaned_COVID_feature$gender)),]
cleaned_COVID_feature$gender <- as.character(cleaned_COVID_feature$gender)
cleaned_COVID_feature$gender <-
 ifelse(cleaned_COVID_feature$gender == "female", 0, 1)
## Fill NA with average age
cleaned_COVID_feature$age <- 
 ifelse(is.na(cleaned_COVID_feature$age), 
 mean(cleaned_COVID_feature$age[!is.na(cleaned_COVID_feature$age)]),
 cleaned_COVID_feature$age)
```

## One-hot encoding

## Collect Symptoms
``` {r}
pasted_symp <- c()
for (val in cleaned_COVID_feature$symptom){
 if (nchar(val) != 0){
 local_vec <- unlist(strsplit(val, ", "))
 for (local_val in local_vec){
 if (is.na(match(local_val, pasted_symp))){
 pasted_symp <- c(pasted_symp, local_val)
 } 
 }
 }
}
pasted_symp <- pasted_symp[-50]

for (index in 1:length(pasted_symp)){
 local_column <- c()
 for (symp in cleaned_COVID_feature$symptom){
 if (grepl(pasted_symp[index], symp)){
 local_column <- c(local_column, 1)
 }
 else{
 local_column <- c(local_column, 0)
 }
 }
 cleaned_COVID_feature <-
 cbind(cleaned_COVID_feature[,1:(2 + index - 1)],
 local_column,
 cleaned_COVID_feature[,
 ((ncol(cleaned_COVID_feature) -
 2):ncol(cleaned_COVID_feature))])
 colnames(cleaned_COVID_feature)[2 + index] <- pasted_symp[index]
}
cleaned_COVID_feature$symptom <- NULL
```

## Combine the recovered and death columnes as response variable, result.

``` {r}
result <- ifelse(cleaned_COVID_feature$death != 0,
 0,
 ifelse(cleaned_COVID_feature$recovered != 0,
 1, 0))
cleaned_COVID_feature <-
 cbind(cleaned_COVID_feature[, 1:(ncol(cleaned_COVID_feature)-2)], Result = result)

head(cleaned_COVID_feature[, 1:10], 3)

colnames(cleaned_COVID_feature)

```

# Feature Selection

``` {r}

model <- glm(Result ~., data = cleaned_COVID_feature, family = 'binomial', control = list(maxit = 50))

intercept_model <- glm(Result ~ 1, data = cleaned_COVID_feature, family = 'binomial', control = list(maxit = 50))


backwards = step(model)


```

