---
title: "Analysis with COVID-19 situation in Canada"
author: "Peter Hou"
date: "03/05/2020"
output:
  pdf_document: defaultp
  html_document: default
---

## Import Necessary Libraries

``` {r}
library(dplyr)
library(readr)
library(ggplot2)
library(reshape2)
```


## Import Dataset

The following datasets are the confirmed, death, cured cases around the world with the date spanning from 2020.1.22 to 2020.4.20

``` {r}
ConfirmedDf <- read.csv("time_series_covid19_confirmed_global.csv")
DeathDf <- read.csv("time_series_covid19_deaths_global.csv")
CuredDf <- read.csv("time_series_covid19_recovered_global.csv")
```

## Examine the Datasets

```{r}
str(ConfirmedDf)
str(DeathDf)
str(CuredDf) 
```
The datasets were all formatted in the same way with only the number's meaning differ with respect to confirmed, death or cured cases.

Note that the cases in each country are separated by province and date. We don't really need the specific detail about each province, we would like to only have the number for each country by date 

Therefore, we would like to consolidate all three datasets for a easier comparison of the trend of COVID-19 cases.

## Data Consolidation

``` {r}
sum_province_by_ctry <- function(CovidDataDf) {
    # Sum all the avaiable number for all provinces in a country by date. 

    aggregate(CovidDataDf[, 3:ncol(CovidDataDf)], by = list(Country = CovidDataDf$Country.Region), sum) # note we use 3:col(CovidDataDf) to only include numeircal columns

}

consolidate_data_by_country <- function(country) {
    ConfirmedDf_country <- ConfirmedDf[which(ConfirmedDf$Country.Region == country), ]
    DeathDf_country <- DeathDf[which(DeathDf$Country.Region == country), ]
    CuredDf_country <- CuredDf[which(CuredDf$Country.Region == country), ]
    # obtaining only the data for the targeted country

    consolidate_country <- list(confirmed = ConfirmedDf_country, death = DeathDf_country, cured = CuredDf_country)
    consolidate_country <- lapply(consolidate_country, sum_province_by_ctry)
    consolidate_country <- rbind(confirmed = consolidate_country$confirmed, death = consolidate_country$death, cured = consolidate_country$cured)
    consolidate_country <- consolidate_country %>% select(-Lat, -Long) # Lat and Long are used to mark the position of provinces on earth, they lost their meaning when we aggregate them. Hence, we remove them from our dataset
n 
    return(consolidate_country)
}

```

Starting from this point, the rest of the report will focus on Canada; but the code can be generalized for any countries. It is just Canada, where I live in, is the target country for analysis in this report.

We now examine the consolidated data for country **Canada** as an example

``` {r}
Canada_consolidate <- consolidate_data_by_country("Canada")

Canada_consolidate[, 1:5]
```

As the data now consolidate in a clear way for comparsion and demonstration. Let's visualize the COVID-19 cases changing trend.

Before the actual visualization work, we need to transpose the dataframe and change the confusing data column format for a better visualization.


``` {r}
t_Canada_consolidate <- t(Canada_consolidate)
head(t_Canada_consolidate)
t_Canada_consolidate <- t_Canada_consolidate[-1, ] # the row of country is not needed
rn <- as.Date(rownames(t_Canada_consolidate), "X%m.%d.%y")
head(t_Canada_consolidate)
t_Canada_consolidate = data.frame(t_Canada_consolidate)
t_Canada_consolidate = cbind(date = rn, t_Canada_consolidate)

rownames(t_Canada_consolidate) <- as.character(as.Date(rownames(t_Canada_consolidate), "X%m.%d.%y"), origin = "2020-01-22")

str(t_Canada_consolidate)

t_Canada_consolidate$confirmed <- as.integer(t_Canada_consolidate$confirmed)
t_Canada_consolidate$death <- as.integer(t_Canada_consolidate$death)
t_Canada_consolidate$cured <- as.integer(t_Canada_consolidate$cured)
```

## Data Visualization

``` {r}
date_for_xaxis <-  c("2020-02-01", "2020-03-01", "2020-04-01")
date_for_xaxis <- as.Date(date_for_xaxis, "%Y-%m-%d")
plot(t_Canada_consolidate$date, t_Canada_consolidate$cured,  main = "COVID-19 Cases in Canada from 2020-01-22 to 2020-04-20", xlab = "Date", ylab = "Number of Cases", ylim = c(0, 40000), xaxt = "n", col = "navyblue", lty = 1, pch = 3)
axis(1, at = t_Canada_consolidate$date[which(t_Canada_consolidate$date %in% date_for_xaxis)], labels = c("Feb", "Mar", "Apr"))
lines(t_Canada_consolidate$date, t_Canada_consolidate$confirmed,  col = "orange", lty = 2, pch = 13)
lines(t_Canada_consolidate$date, t_Canada_consolidate$death,  col = "red", lty = 3, pch = 16)
legend("topleft", col = c("navyblue", "orange", "red"),
       legend = c("cured", "confirmed", "death"),
       pch  = c(3, 13, 16), lty = c(1, 2, 3))

                                        # Below is the visualization done by ggplot, only for pratice purpose
reshaped_consolidate_data <- melt(t_Canada_consolidate, id.vars="date")

head(reshaped_consolidate_data,3)

p1 <- ggplot(reshaped_consolidate_data, aes(date, value, col=variable)) +
    geom_point() +  ggtitle("Canada COVID-19 Record") + labs(y = "Cases Number", x = "Date", color = "Case Type") 

p1

```

You may be confused with the x-axis labels on my ggplot output. That is Chinese characters. Unfortunately, my laptop has come with the initial configuration of Chinese character which followed by ggplot. I have tried numerous ways to change it to English characters but failed all the time. **If you know anything about a feasible solution, please don't hesitate to contact me**

## Statistical Inference
Reading from the graph, Covid-19 confirmed cases drastically rose from January to March, reached 30000 in April and showed no sign of a decreasing afterwards. Many articles I have read commonly would compare the COVID-19 with other respiratory virus, such as SARS, in terms of their spead or transmission ability in the human society. 

The assessment of COVID-19 transimision ability will be the focus of the remaining report.

### SIR Model
To find the transmission ability of COVID-19, we will utilize the SIR model for epidemic. This model simplifies complex situation of a epidemic (pandemic in this case) into a single system where individuals can only have three states (considered as function of time t):

1. Susceptible: not infected yet and not immune, denoted by S
2. Infected: infected and can infect others, denote by I
3. Removed: immuned to the disease or cannot infect others, denote by R

#### Futher Rate Assumption
In this model, we assume further there are 3 more quantities: \par
1. $\lambda$ is the uniform probability of being infected, note that it is not constant. It must consider other factors like the spreadness of the virus. E.g. if everyone around you has been infected, then there is a larger change that you get infected than an average person. \par
2. $\mu$ is the uniform probability of being dead. This is a very small quantity since it is the natural death rate of the whole world population. We ignore this quantity. \par
3. $\gamma$ is the uniform probability of being recovered \par

Based on the model assumption and the further rate assumption, we were able to obtain the following system of Diffferential Equations

$$
  \begin{aligned}
    \frac{dS}{dt} &= B - \lambda S - \mu S\\
    \frac{dI}{dt} &= \lambda S - \gamma I - \mu I\\
    \frac{dR}{dt} &= \gamma I - \mu R
  \end{aligned}
$$
where $B$ is the bith rate of the population

#### Relationship between each assumed varible

By solving the system of differential equation, we were able to obtain
\[I(t) = I_{0}e^{(R_{0} - 1)(\gamma + \mu)t} \approx I_{0}e^{(R_{0}
    - 1) \gamma t}\]
Where $I_{0}$ is the intial amount of people getting infected. Note
that clearly not every quantity is constant otherwise $I(t)$ will never
decrease. This is a model for early stage

**$R_{0}$** is usually what we call **reproduction number**.

#### Goal Setting
The goal here for the modeling using SIR model is to **estimate the reproduction number for COVID-19**, which we can use to compare with similar virus as mentioned above like SARS.

#### Estimation of $\gamma$
From the given fomula above, we have observation for $I(t)$ at different time point t, from january 2020 to April 2020

With our assumption, $\gamma$ is the rate of recovery of the infected population **over a period**. Hence, $1/\gamma$ is the average time length of the period to get one recovered. We can model this process using a Poisson ditribution with $Y_i \sim Poi(\gamma)$ and we need to estimate $\gamma$. In this case, $Y_i$ is the number of recovered people over a fixed period of time.

As $\gamma$ is dealing with recovery cases, then our sample in this case would be 

``` {r}
recover_sample <- t_Canada_consolidate$cured
recover_sample <- diff(recover_sample, lag = 1) 
# This is to get the data fro difference between any two days since cured is the total cases and we want each days' recovery data.


```
Now we perform **Maximum Likelihood Method** and get



  \begin{equation}
  \begin{aligned}
    L(\theta) &=
    \Pi_{i=1}^{n}\frac{e^{(-\theta)}\theta^{y_{i}}}{y_{i}!}\\
              &=
              (\Pi_{i=1}^{n}\frac{1}{y_{i}!})e^{-n\theta}\theta^{\sum_{i=1}^{n}y_{i}}\\
              &=Ce^{-n\theta}\theta^{\sum_{i=1}^{n}y_{i}}
\end{aligned}
  \end{equation}
where the C is standing for the proportionality.

We write this function in R
``` {r}
poiLF <- function(theta) {exp(-length(recover_sample) * theta) * theta ^ sum(recover_sample)}

```

We can directly graph poiLF to find out which theta maximizes this function, however,

``` {r}
sum(recover_sample)

```

The value is too big, it's going to be very difficult to visualize the likelihood function on this data and find out the point that maximizes the function. Hence, we use the loglikelihood function.

``` {r}
poiLLF <- function(theta) {
    -length(recover_sample) * theta+sum(recover_sample) * log(theta)
}

curve(poiLLF, xlim = c(0, 300), xlab = "theta", main = "Log-Likelihood Function")
```

From the graph, we can roughly get that the theta that maximizes the function is get around 150.

``` {r}
mle_result <- optimize(poiLLF, interval = c(0, 300),maximum = TRUE)
mle_result
```

From our output, 142.5341 is the maximum likelihood estimate of gamma. 
``` {r}
gamma_mle <- mle_result$maximum
```

#### Estimation of $R_0$
\[  I(t) = I_{0}e^{(R_{0} - 1)(\gamma + \mu)t} \approx I_{0}e^{(R_{0}
    - 1) \gamma t}  \]

The relationship of $R_0$ in this original equation may be a bit hard to model. However, if we take a log at both sides of equation, we get

\[ log I = log I_0 + (R_0 - 1) \gamma t\]

It demonstrates a linear relationship between log I and the time t. Hence, we can apply simple lienar regression for modelling 


As before, we need data for I, the confirmed_sample, but we do not need daily new increase. 
``` {r}
# log is not defined for log(0), hence, we remove the points where the confirmed cases are of 0

confirmed_sample <- t_Canada_consolidate$confirmed[t_Canada_consolidate$confirmed != 0]
log_confirmed_sample <- log(confirmed_sample)
logconfirmed_df <- data.frame(Days = 1:length(log_confirmed_sample),
                              Log_Confirmed = log_confirmed_sample)

ggplot(logconfirmed_df, aes(x = Days, y = Log_Confirmed)) + geom_point() + geom_smooth(method = "lm") + ggtitle("Log-Transformed Confirmed Cases in Canada")
```
``` {r}
log_lm <- lm(Log_Confirmed ~ Days, data = logconfirmed_df)
summary(log_lm)
```

From the regression coefficient, we estimated that `logI \sim -0.698895 + 0.135525t`. Thus, $(R_0 - 1)\gamma = 0.135525$. We have estimated that $\hat{\gamma} = 142.5341$. Hence, 
\[ \hat{R_0} = \frac{0.135525}{\gamma} + 1 = 1.000951]

The number 1 is what separats a highly contagious disease(>1) and a lowly contagious disease (<1). For SARS-Cov, it has a reproductive number of 2-3. Note, COVID-19 is SARS-Cov2. In comparison, based on our assumption and result, COVID-19 is less infectious than SARS-Cov. However, this is certainly not the case with what happens in the real world, which the confirmed cases are exponentially increasing everyday and every country is issuing serious warning to every citizen. 

The possible cause for the low reproductive number is likely due to the equal weight on every datapoint in the dataset which the first two months of small confirmed number averaged out the last two months of scary and exponential increase.

In the meanwhile, the confirmed cases leveled around 0 in the first two months does not mean there are only this many confirmed cases in Canada at that time. It only means that this many of confirmed cases have chosen to take a test and found out they are positive. The initial symptom for youngsters are mild; which many people may not take a serious attention. 



```{r}

patient_record <- read_csv("COVID19_line_list_data.csv")

```











